---
title: "R_Markdown_Valerie"
author: "Valerie Yu"
date: "`r Sys.Date()`"
output: pdf_document
---

# Valerie's R Assignment
Setting up the R Markdown and loading in R library.
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

rm(list = ls())
library(tidyverse)
library(ggplot2)
library(tidyr)
```


## Part I
### _Data Inspection_

```{r}
# Loading in data from "data" folder and looking at column names. 
genotypes <- read_tsv("data/fang_et_al_genotypes.txt")
colnames(genotypes)[1:15]
snp_position <- read_tsv("data/snp_position.txt")
colnames(snp_position)[1:15]


# Inspecting the data
dim(genotypes) # 2782 rows  986 columns
names(genotypes) # variable names 
head(genotypes) # 1st Sample_ID, 3rd Group
table(genotypes$Group) 
  # maize count: ZMMIL 290, ZMMLR 1256, and ZMMMR 27 = 1573
  # teosinte count: ZMPBA 900, ZMPIL 41, and ZMPJA 34 = 975


dim(snp_position) # 983 rows  15 columns
names(snp_position) # variable names
head(snp_position) # 1st SNP_ID, 3rd Chromosome, 4th Position, etc.
table(snp_position$Chromosome) # 6 multiple and 27 unknown
```
`fang_et_al_genotypes.txt`
* There are 2782 rows and 986 columns.
* The first 3 columns are `Sample_ID`, `JG_OTU`, and `Group`. Then the rest are SNP data (983).

`snp_position.txt`
* There are 983 rows and 15 columns.
* The columns are named `SNP_ID`, `Chromosome`, `Position`, etc.

### _Data Processing_
`fang_et_al_genotypes.txt` 
First, I am cutting out the groups I need for maize and teosinte.
```{r}
# Filtering `fang_et_al_genotypes.txt` 
maize_genotypes <- genotypes %>% filter(Group == "ZMMIL" |
                                        Group == "ZMMLR" |
                                        Group == "ZMMMR" )
teosinte_genotypes <- genotypes %>% filter(Group == "ZMPBA" |
                                           Group == "ZMPIL" |
                                           Group == "ZMPJA" )
# Checking: both result is 3, which is accurate
n_distinct(maize_genotypes$Group)
n_distinct(teosinte_genotypes$Group) 
```

Then I will transpose the genotype files so that columns become rows by using t() before joining.
```{r}
# MAIZE
t_maize_genotypes <- as.data.frame(t(maize_genotypes))
t_maize_genotypes <- t_maize_genotypes[-c(1,2,3),]
dim(t_maize_genotypes) # 983 rows just like snp_position

# TEOSINTE
t_teosinte_genotypes <- as.data.frame(t(teosinte_genotypes))
t_teosinte_genotypes <- t_teosinte_genotypes[-c(1,2,3),]
dim(t_teosinte_genotypes) # 983 rows just like snp_position
```

Now I am joining snp_position to both of these genotype files.
```{r}
# MAIZE
joined_maize

# TEOSINTE
joined_teosinte
```




For maize (Group = ZMMIL, ZMMLR, and ZMMMR in the third column of the fang_et_al_genotypes.txt file) we want 20 files in total:
```{r}

```

For teosinte (Group = ZMPBA, ZMPIL, and ZMPJA in the third column of the fang_et_al_genotypes.txt file) we want 20 files in total:
```{r}

```

## Part II Visualization

### _SNPs per chromosome_
What is the distribution of SNPs on and across chromosomes?
Are there more SNP positions in maize or teosinte individuals?

### _Missing data and amount of heterozygosity_
What is the proportion of homozygous and heterozygous sites as well as missing data in each sample and each group?  

### _Your own visualization_
Visualize one other feature of the dataset. The choice is up to you!